---
title: "Report 2 - Two-Way ANOVA for Initial Response Time"
author: "Mavericks"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# Introduction

This analysis tests whether Initial Response Time (IRT) differs as a function of:

- **Platform**: Lab vs Phone  
- **Condition**: Single vs Multiple Target  
- **Interaction**: Whether the effect of condition depends on platform  

A two-way ANOVA is conducted on **log-transformed IRT** because reaction time data are positively skewed.

---

```{r}
#====================================================
# LOAD PACKAGES
#====================================================
library(dplyr)
library(ggplot2)

#====================================================
# FILE PATHS
#====================================================
path_multi_lab   <- "/home/kritika/MTech/Sem4/BRSM/Project/Attention Task Validation/data_brsm/multiple/lab"
path_multi_phone <- "/home/kritika/MTech/Sem4/BRSM/Project/Attention Task Validation/data_brsm/multiple/phone"
path_single_lab  <- "/home/kritika/MTech/Sem4/BRSM/Project/Attention Task Validation/data_brsm/single/lab"
path_single_phone<- "/home/kritika/MTech/Sem4/BRSM/Project/Attention Task Validation/data_brsm/single/phone"

#====================================================
# READ ALL CSV FILES FROM FOLDER
#====================================================
read_folder_csv <- function(folder){

  files <- list.files(folder, pattern="\\.csv$", full.names=TRUE)

  data_list <- lapply(files, function(f){

    read.csv(
      f,
      stringsAsFactors = FALSE,
      colClasses = "character",
      check.names = FALSE
    )

  })

  bind_rows(data_list)
}

#====================================================
# EXTRACT LAB IRT
# mouse.time contains list of click times
# Use first value and convert sec -> ms
#====================================================
extract_lab_irt <- function(df){

  cols <- names(df)

  idx <- grep("mouse.time", cols, ignore.case = TRUE)

  if(length(idx) == 0){
    stop("mouse.time column not found in lab data")
  }

  time_col <- cols[idx[1]]

  cat("Using lab column:", time_col, "\n")

  df %>%
    mutate(
      first_time = sub("^\\[\\s*([0-9.]+).*$", "\\1", .data[[time_col]]),
      IRT = suppressWarnings(as.numeric(first_time)) * 1000
    ) %>%
    filter(!is.na(IRT), IRT > 0) %>%
    select(IRT)
}

#====================================================
# EXTRACT PHONE IRT
# Detect InitialResponseTime column robustly
#====================================================
extract_phone_irt <- function(df){

  cols <- names(df)

  idx <- grep("Initial.*Response.*Time", cols, ignore.case = TRUE)

  if(length(idx) == 0){
    print(cols)
    stop("Phone IRT column not found")
  }

  time_col <- cols[idx[1]]

  cat("Using phone column:", time_col, "\n")

  df %>%
    mutate(
      IRT = suppressWarnings(as.numeric(.data[[time_col]]))
    ) %>%
    filter(!is.na(IRT), IRT > 0) %>%
    select(IRT)
}

#====================================================
# READ + CLEAN ALL FOUR GROUPS
#====================================================

# Multiple - Lab
multi_lab_raw <- read_folder_csv(path_multi_lab)

multi_lab <- extract_lab_irt(multi_lab_raw) %>%
  mutate(
    Platform = "Lab",
    Condition = "Multiple"
  )

# Multiple - Phone
multi_phone_raw <- read_folder_csv(path_multi_phone)

multi_phone <- extract_phone_irt(multi_phone_raw) %>%
  mutate(
    Platform = "Phone",
    Condition = "Multiple"
  )

# Single - Lab
single_lab_raw <- read_folder_csv(path_single_lab)

single_lab <- extract_lab_irt(single_lab_raw) %>%
  mutate(
    Platform = "Lab",
    Condition = "Single"
  )

# Single - Phone
single_phone_raw <- read_folder_csv(path_single_phone)

single_phone <- extract_phone_irt(single_phone_raw) %>%
  mutate(
    Platform = "Phone",
    Condition = "Single"
  )

#====================================================
# COMBINE DATA
#====================================================
df <- bind_rows(
  multi_lab,
  multi_phone,
  single_lab,
  single_phone
)

#====================================================
# FINAL CLEANING
#====================================================
df <- df %>%
  filter(!is.na(IRT), IRT > 0) %>%
  mutate(
    Platform = factor(Platform),
    Condition = factor(Condition),
    logIRT = log(IRT)
  )

#====================================================
# CHECK COUNTS
#====================================================
table(df$Platform, df$Condition)

head(df)
summary(df)
```


```{r}
df <- bind_rows(multi_lab, multi_phone, single_lab, single_phone)

df <- df %>%
  filter(!is.na(IRT), IRT > 0) %>%
  mutate(
    Platform = factor(Platform),
    Condition = factor(Condition),
    logIRT = log(IRT)
  )

head(df)
summary(df)
```



```{r}
table(df$Platform, df$Condition)
```
Visualisation
```{r}
ggplot(df, aes(x=Condition, y=IRT, fill=Platform)) +
  geom_boxplot(alpha=.7) +
  facet_wrap(~Platform) +
  labs(
    title="Initial Response Time by Platform and Condition",
    y="IRT (ms)",
    x="Condition"
  ) +
  theme_minimal()
```


One-Way ANOVA

```{r}
#=========================================
# ONE-WAY ANOVA ACROSS FOUR GROUPS
#=========================================

# Create combined 4-level group variable
df <- df %>%
  mutate(
    Group = interaction(Platform, Condition, sep = "_")
  )

# Check counts
table(df$Group)

# Run one-way ANOVA
oneway_model <- aov(logIRT ~ Group, data = df)

# Results
summary(oneway_model)
```
A one-way ANOVA was conducted to compare log-transformed Initial Response Time across the four experimental groups (Lab-Multiple, Phone-Multiple, Lab-Single, Phone-Single). The analysis revealed a significant difference among groups, F(3, 1078) = 92.39, p < .001, indicating that response time varied across experimental conditions. To further examine whether these differences were attributable to Platform, Condition, and their interaction, a two-way factorial ANOVA was subsequently conducted.

```{r}
TukeyHSD(oneway_model)
```
Tukey post hoc comparisons indicated that the Phone-Single group had significantly higher response times than Lab-Multiple (p < .001), Phone-Multiple (p < .001), and Lab-Single (p < .001). No significant differences were observed among Lab-Multiple, Lab-Single, and Phone-Multiple conditions. These findings suggest that the elevated response times were primarily driven by the Phone-Single condition.


Two-Way ANOVA
```{r}
anova_model <- aov(logIRT ~ Platform * Condition, data=df)
summary(anova_model)
```

Effect Size (Eta Squared)
```{r}
anova_tab <- anova(anova_model)

SS_total <- sum(anova_tab$"Sum Sq")

eta_sq <- data.frame(
  Effect = rownames(anova_tab),
  SS = anova_tab$"Sum Sq",
  Eta_Squared = anova_tab$"Sum Sq" / SS_total
) %>%
  filter(Effect != "Residuals")

eta_sq
```

Effect Size Interpretation
```{r}
eta_sq <- eta_sq %>%
  mutate(
    Magnitude = case_when(
      Eta_Squared >= 0.14 ~ "Large",
      Eta_Squared >= 0.06 ~ "Medium",
      Eta_Squared >= 0.01 ~ "Small",
      TRUE ~ "Negligible"
    )
  )

eta_sq
```

```{r}
png("interaction_plot.png", width=1000, height=1000, res=300)

interaction.plot(
  x.factor = df$Condition,
  trace.factor = df$Platform,
  response = df$logIRT,
  fun = mean,
  type = "b",
  pch = c(19,17),
  col = c("blue","red"),
  xlab = "Condition",
  ylab = "Mean log(IRT)",
  trace.label = "Platform",
  main = "Interaction Plot"
)

dev.off()

# Show in RStudio
interaction.plot(
  x.factor = df$Condition,
  trace.factor = df$Platform,
  response = df$logIRT,
  fun = mean,
  type = "b",
  pch = c(19,17),
  col = c("blue","red"),
  xlab = "Condition",
  ylab = "Mean log(IRT)",
  trace.label = "Platform",
  main = "Interaction Plot"
)
```

Assumption Check: Residual Normality
```{r}
res <- residuals(anova_model)
shapiro.test(sample(res, min(length(res), 5000)))
```

Assumption Check: Homogeneity of Variance
```{r}
bartlett.test(logIRT ~ interaction(Platform, Condition), data=df)
```
Interpretation of Results:

1. Significant Platform Effect:
   A significant main effect of Platform was observed, F(1, 1078) = 163.10, p < .001, eta² = .120.
   This indicates that response times differed significantly between the Lab and Phone platforms.
   The medium effect size suggests that platform type had the strongest practical influence on performance.

2. Significant Condition Effect:
   A significant main effect of Condition was found, F(1, 1078) = 52.01, p < .001, eta² = .038.
   This indicates that response times differed significantly between the Single and Multiple target tasks.
   The effect size was small, suggesting a modest but meaningful effect of task condition.

3. Significant Interaction Effect:
   A significant Platform × Condition interaction was found, F(1, 1078) = 62.07, p < .001, eta² = .046.
   This indicates that the effect of task condition depended on the platform used.
   In other words, the difference between Single and Multiple tasks was not the same for Lab and Phone conditions.

4. Effect Size Benchmarks:
   eta² = .01  Small Effect
   eta² = .06  Medium Effect
   eta² = .14  Large Effect

5. Assumption Checks:
   Shapiro-Wilk and Bartlett tests were significant, indicating deviations from normality
   and equal variances. However, given the large sample size and use of log-transformed
   response times, ANOVA was considered sufficiently robust.

6. Overall Conclusion:
   Response time performance was influenced by platform type, task condition,
   and their interaction. Platform differences showed the largest effect,
   supporting the importance of device context in attention task performance.


Non-parametric Robustness Check: Kruskal–Wallis Test
```{r}
#=========================================
# KRUSKAL-WALLIS TEST ACROSS FOUR GROUPS
#=========================================

# Ensure group variable exists
df <- df %>%
  mutate(
    Group = interaction(Platform, Condition, sep = "_")
  )

# Run Kruskal-Wallis test on raw IRT
kw_model <- kruskal.test(IRT ~ Group, data = df)

kw_model
```
Because assumption checks indicated deviations from normality and homogeneity of variance, a supplementary Kruskal–Wallis test was conducted across the four experimental groups. The test was significant, $\chi^2$(3) = 168.1, p < .001, indicating that response time distributions differed across groups. This result is consistent with the ANOVA findings and supports the robustness of the observed group differences.

```{r}
pairwise.wilcox.test(
  df$IRT,
  df$Group,
  p.adjust.method = "bonferroni"
)
```
Pairwise Wilcoxon rank-sum tests with Bonferroni correction showed that the Phone-Single group differed significantly from Lab-Multiple, Phone-Multiple, and Lab-Single conditions (all p < .001). No significant differences were observed among Lab-Multiple, Lab-Single, and Phone-Multiple groups (all adjusted p = 1.00). These findings mirror the Tukey post hoc results and further support the robustness of the observed group pattern.


