---
title: "BRSM Report 1 - Data Visualization"
output:
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
```

```{r load-cleaned}
data_list <- readRDS("cleaned_data.rds")

lab_multi_clean    <- data_list$lab_multi_clean
lab_single_clean   <- data_list$lab_single_clean
phone_multi_clean  <- data_list$phone_multi_clean
phone_single_clean <- data_list$phone_single_clean

# Ensure all key columns are numeric
lab_multi_clean$irt_ms  <- as.numeric(lab_multi_clean$irt_ms)
lab_single_clean$irt_ms <- as.numeric(lab_single_clean$irt_ms)

phone_multi_clean  <- phone_multi_clean  %>%
  mutate(across(c(irt_ms, false_alarms, hit_rate, avg_inter_ms, Level), as.numeric))
phone_single_clean <- phone_single_clean %>%
  mutate(across(c(irt_ms, false_alarms, hit_rate, avg_inter_ms, Level), as.numeric))

# Combined dataset: all 4 groups for IRT figures
combined_irt <- bind_rows(
  lab_single_clean   %>% select(participant_id, irt_ms) %>%
    mutate(Group = "Lab\nSingle",    Platform = "Lab",   Condition = "Single"),
  lab_multi_clean    %>% select(participant_id, irt_ms) %>%
    mutate(Group = "Lab\nMultiple",  Platform = "Lab",   Condition = "Multiple"),
  phone_single_clean %>% select(participant_id, irt_ms) %>%
    mutate(Group = "Phone\nSingle",  Platform = "Phone", Condition = "Single"),
  phone_multi_clean  %>% select(participant_id, irt_ms) %>%
    mutate(Group = "Phone\nMultiple",Platform = "Phone", Condition = "Multiple")
) %>% filter(!is.na(irt_ms))

# Combined phone dataset 
phone_combined <- bind_rows(
  phone_single_clean %>% mutate(Condition = "Single"),
  phone_multi_clean  %>% mutate(Condition = "Multiple")
)
```

```{r theme}
theme_brsm <- theme_classic() +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, color = "grey40"),
    axis.title      = element_text(size = 11),
    strip.text      = element_text(face = "bold"),
    legend.position = "bottom"
  )

dir.create("figures", showWarnings = FALSE)
```


# Figure 1 — IRT Distribution Shape (Histograms, All 4 Groups)

```{r fig1, fig.height=6}
if (!dir.exists("figures")) {
  dir.create("figures")
}
p1 <- ggplot(combined_irt, aes(x = irt_ms, fill = Group)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.85) +
  facet_wrap(~Group, scales = "free") +
  labs(
    title = "Distribution of Initial Response Time (IRT) by Group",
    x     = "Initial Response Time (ms)",
    y     = "Count"
  ) +
  theme_brsm + theme(legend.position = "none")

p1
ggsave("figures/fig1_irt_histogram.jpeg", plot = p1,
       width = 9, height = 6, dpi = 150)
```

**Interpretation:** All four groups show right-skewed IRT distributions, most responses
are fast but a small number of trials produce very long latencies. This is typical of
reaction time data and justifies using the median as the measure of central tendency.
Phone-Single shows the widest spread (up to 14,711 ms), while both lab conditions are
tightly concentrated between 800-2,000 ms, reflecting the controlled laboratory
environment.

---

# Figure 2 — IRT Spread and Outliers (Box Plots, All 4 Groups)

```{r fig2}
p2 <- ggplot(combined_irt, aes(x = Group, y = irt_ms, fill = Group)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 2, alpha = 0.8) +
  labs(
    title = "IRT Spread and Outliers by Group",
    x     = "Group",
    y     = "Initial Response Time (ms)"
  ) +
  theme_brsm + theme(legend.position = "none")

p2
ggsave("figures/fig2_irt_boxplot.jpeg", plot = p2,
       width = 7, height = 5, dpi = 150)
```

**Interpretation:** Lab-Single (1,440 ms) and Lab-Multiple (1,419 ms) medians are nearly
identical, target number didn't substantially change central IRT in the lab.
Phone-Multiple is similarly compact (median ~1,449 ms), but Phone-Single stands out with
a higher median (~2,028 ms) and far more extreme outliers, suggesting that finding a lone
target on a small touchscreen is sporadically very difficult.

---

# Figure 3 — Full IRT Distribution by Platform and Condition (Violin + Jitter)

```{r fig3}
p3 <- ggplot(combined_irt, aes(x = Condition, y = irt_ms, fill = Platform)) +
  geom_violin(trim = FALSE, alpha = 0.6, position = position_dodge(0.8)) +
  geom_jitter(aes(color = Platform),
              position = position_jitterdodge(0.2), size = 0.8, alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               position = position_dodge(0.8), color = "black", fill = "white") +
  labs(
    title    = "IRT Full Distribution by Platform and Condition",
    subtitle = "Diamond = Mean; dots = individual trials/sessions",
    x        = "Condition",
    y        = "Initial Response Time (ms)"
  ) +
  theme_brsm

p3
ggsave("figures/fig3_irt_violin.jpeg", plot = p3,
       width = 8, height = 5, dpi = 150)
```

**Interpretation:** Lab distributions (red) are narrow and similarly shaped in both
conditions. The phone single-target violin (teal, right panel) has a dramatically wider
body with a long upper tail, while the phone multiple-target violin is comparatively
compact. This reveals a platform × condition interaction: high variability is specific to
phone + single-target, not to the phone platform in general. The mean (diamond) lies
above the median in all groups, confirming right skew.

---

# Figure 4 — False Alarm Distribution by Condition (Phone)

```{r fig4}
p4 <- ggplot(phone_combined, aes(x = false_alarms, fill = Condition)) +
  geom_histogram(binwidth = 1, position = "dodge",
                 color = "white", alpha = 0.85) +
  labs(
    title = "Distribution of False Alarms by Condition (Phone)",
    x     = "Number of False Alarms",
    y     = "Count"
  ) +
  theme_brsm

p4
ggsave("figures/fig4_fa_histogram.jpeg", plot = p4,
       width = 7, height = 5, dpi = 150)
```

**Interpretation:** Both conditions show strongly zero-inflated false alarm distributions,
the overwhelming majority of sessions produced zero false alarms. The multiple-target
condition has a heavier right tail (mean = 0.61 vs 0.21 for single-target), with more
sessions recording 1–5 false alarms. This partially supports H4: while false alarms are
rare overall, they are more frequent under higher attentional load.

